\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage[hidelinks]{hyperref}

\title{Algorithms for Approximate Inversion of Weave Structures}
% \author{Peter E. Francis}
\date{\today}

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{observation}{Observation}
\newtheorem{proposition}{Proposition}
\newtheorem{example}{Example}

\begin{document}

\maketitle

\begin{abstract}
These notes study the inverse problem for shaft-loom weaving: given a desired
binary drawdown, find a threading, tie-up, and treadling that reproduce it
exactly when possible, and approximate it well when exact reproduction is
impossible under loom constraints.
\end{abstract}

\section{Introduction}

A single-layer woven structure may be idealized as a binary matrix
\[
  \widehat{D} \in \{0,1\}^{P \times W},
\]
where \(P\) is the number of weft picks and \(W\) is the number of warp ends.
Rows index picks and columns index warp ends. We use the convention
\[
  \widehat{D}_{ij}
  =
  \begin{cases}
    1, & \text{warp end } j \text{ passes over weft pick } i,\\
    0, & \text{weft pick } i \text{ passes over warp end } j.
  \end{cases}
\]
In this language, a Jacquard loom is essentially unconstrained: each warp end
may be controlled independently on each pick, so every finite binary matrix is
attainable, ignoring physical considerations such as yarn behavior, abrasion,
float length, and cloth stability. Matrix and algebraic models of loom
constraints go back at least to Lourie's work on loom-constrained designs, which
represents a weave as a binary matrix and studies how to recover loom data from
that matrix~\cite{lourie1969}.

The situation is very different for a shaft loom. A warp end is threaded through
one shaft, and all warp ends on the same shaft move together. Thus a shaft loom
cannot directly control individual warp ends; it controls equivalence classes of
warp ends. Let \(T\) be the number of treadles, \(S\) the number of shafts, and
\(R\) the maximum number of treadles allowed to be pressed on a single pick.
We encode a draft by three binary matrices:
\[
  A \in \{0,1\}^{P \times T},
  \qquad
  B \in \{0,1\}^{S \times T},
  \qquad
  C \in \{0,1\}^{S \times W}.
\]
Here \(A\) is the treadling matrix, with \(A_{it}=1\) if treadle \(t\) is pressed
on pick \(i\); \(B\) is the tie-up matrix, with \(B_{st}=1\) if treadle \(t\)
lifts shaft \(s\); and \(C\) is the threading matrix, with \(C_{sj}=1\) if warp
end \(j\) is threaded on shaft \(s\). Each column of \(C\) contains exactly one
nonzero entry, and each row of \(A\) contains at most \(R\) nonzero entries.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
    x=0.54cm,
    y=0.54cm,
    gridline/.style={draw=black!22, line width=0.25pt},
    sectionline/.style={draw=black!70, line width=0.7pt},
    block/.style={fill=black!86},
    label/.style={font=\small\bfseries}
  ]
    \def\gap{0.45}
    \newcommand{\sectiongrid}[2]{%
      \foreach \x in {0,...,#1} {\draw[gridline] (\x,0) -- (\x,#2);}
      \foreach \y in {0,...,#2} {\draw[gridline] (0,\y) -- (#1,\y);}
      \draw[sectionline] (0,0) rectangle (#1,#2);
    }

    \begin{scope}[shift={(0,12+\gap)}]
      \foreach \x/\y in {0/3,1/2,2/1,3/0,4/1,5/2,6/3,7/2,8/1,9/0,10/1,11/2} {
        \fill[block] (\x,\y) rectangle ++(1,1);
      }
      \sectiongrid{12}{4}
      \node[label, above] at (6,4.45) {threading \(C\)};
    \end{scope}

    \begin{scope}[shift={(12+\gap,12+\gap)}]
      \foreach \x/\y in {0/3,0/2,1/2,1/1,2/1,2/0,3/0,3/3,4/3,4/1,5/2,5/0} {
        \fill[block] (\x,\y) rectangle ++(1,1);
      }
      \sectiongrid{6}{4}
      \node[label, above] at (3,4.45) {tie-up \(B\)};
    \end{scope}

    \begin{scope}
      \foreach \x/\y in {
        0/11,1/11,5/11,6/11,7/11,11/11,
        1/10,2/10,4/10,5/10,7/10,8/10,10/10,11/10,
        2/9,3/9,4/9,8/9,9/9,10/9,
        0/8,3/8,6/8,9/8,
        0/7,2/7,4/7,6/7,8/7,10/7,
        1/6,3/6,5/6,7/6,9/6,11/6,
        0/5,1/5,5/5,6/5,7/5,11/5,
        0/4,3/4,6/4,9/4,
        1/3,2/3,4/3,5/3,7/3,8/3,10/3,11/3,
        0/2,2/2,4/2,6/2,8/2,10/2,
        2/1,3/1,4/1,8/1,9/1,10/1,
        1/0,3/0,5/0,7/0,9/0,11/0
      } {
        \fill[block] (\x,\y) rectangle ++(1,1);
      }
      \sectiongrid{12}{12}
      \node[label, below] at (6,-0.45) {drawdown \(\widehat{D}\)};
    \end{scope}

    \begin{scope}[shift={(12+\gap,0)}]
      \foreach \x/\y in {0/11,1/10,2/9,3/8,4/7,5/6,0/5,3/4,1/3,4/2,2/1,5/0} {
        \fill[block] (\x,\y) rectangle ++(1,1);
      }
      \sectiongrid{6}{12}
      \node[label, below] at (3,-0.45) {treadling \(A\)};
    \end{scope}
  \end{tikzpicture}
  \caption{A draft written as matrices.  Dark cells are entries equal to \(1\).
  This example uses \(W=P=12\), \(S=4\), and \(T=6\): the threading \(C\),
  tie-up \(B\), and treadling \(A\) determine the drawdown by thresholding the
  redundancy matrix \(D=AB^TC\).}
  \label{fig:example-draft}
\end{figure}

The natural product in this notation is
\[
  D = A B^{T} C.
\]
The matrix \(D\) is not merely binary. It records the redundancy with which each
cell is activated: \(D_{ij}=0\) means that warp end \(j\) is not lifted over pick
\(i\), \(D_{ij}=1\) means that it is lifted by one pressed treadle, and
\(D_{ij}=2\) means that two pressed treadles both lift the shaft carrying warp
end \(j\), and so on. The binary drawdown associated to the draft is therefore
\[
  \widehat{D}_{ij}
  =
  \begin{cases}
    0, & D_{ij}=0,\\
    1, & D_{ij}>0.
  \end{cases}
\]
This is not ordinary unconstrained matrix factorization; it is a discrete
factorization through treadle, shaft, and warp-thread incidence matrices.

\begin{proposition}[Finite shaft-loom attainability with multiple treadles]
Let \(E\in\{0,1\}^{P\times W}\) be a finite binary drawdown. Delete duplicate
columns of \(E\), and call the resulting matrix \(E_{\mathrm{col}}\). Let
\(S_0\) be the number of columns of \(E_{\mathrm{col}}\), and let
\(\mathcal{R}\subseteq\{0,1\}^{S_0}\) be the set of distinct rows of
\(E_{\mathrm{col}}\).

For a set \(\mathcal{B}\subseteq\{0,1\}^{S_0}\), define its
\(R\)-fold union closure by
\[
  \mathcal{U}_R(\mathcal{B})
  =
  \left\{
    b_1\vee b_2\vee\cdots\vee b_k
    :
    0\le k\le R,\ b_1,\dots,b_k\in\mathcal{B}
  \right\},
\]
where \(\vee\) denotes coordinatewise Boolean OR and the empty union is the
zero vector. Then \(E\) is attainable on a shaft loom with at most \(S\) shafts,
\(T\) treadles, and at most \(R\) treadles pressed on any pick if and only if
\[
  S_0\le S
  \qquad\text{and}\qquad
  \mathcal{R}\subseteq \mathcal{U}_R(\mathcal{B})
\]
for some set \(\mathcal{B}\subseteq\{0,1\}^{S_0}\) with
\(|\mathcal{B}|\le T\).

When \(R=1\) and each pick is required to use a single treadle, this reduces to
the simpler criterion that the reduced drawdown has at most \(S\) distinct
column types and at most \(T\) distinct row types. With enough treadles or with a
dobby-style liftplan, the minimum number of shafts is the number of distinct
warp-column behaviors. For an infinite periodic prefabric, the same criterion
applies to a fundamental period block.
\end{proposition}

\begin{proof}
First suppose that \(E\) is produced by a shaft loom with at most \(S\) shafts,
\(T\) treadles, and at most \(R\) treadles pressed on a pick. If two warp ends
are threaded on the same shaft, then they are lifted on exactly the same picks,
so their columns in \(E\) are identical. Therefore the number of distinct column
behaviors is at most the number of shafts, hence \(S_0\le S\).

After duplicate columns are removed, each treadle determines a binary vector in
\(\{0,1\}^{S_0}\): its entries record which of the \(S_0\) column types are
lifted by that treadle. Let \(\mathcal{B}\) be the set of these treadle vectors.
Then \(|\mathcal{B}|\le T\). If pick \(i\) presses treadles
\(t_1,\dots,t_k\), with \(k\le R\), then its reduced row is the Boolean union
of the corresponding treadle vectors,
\[
  b_{t_1}\vee\cdots\vee b_{t_k}.
\]
Thus every distinct reduced row lies in \(\mathcal{U}_R(\mathcal{B})\).

Conversely, suppose \(S_0\le S\) and
\(\mathcal{R}\subseteq\mathcal{U}_R(\mathcal{B})\) for some
\(\mathcal{B}=\{b_1,\dots,b_m\}\subseteq\{0,1\}^{S_0}\) with \(m\le T\).
Assign each distinct column type to one shaft, threading every warp end with
that column behavior through the corresponding shaft. Assign the vector
\(b_t\) to treadle \(t\), tying it to exactly those shafts whose column types
have value \(1\) in \(b_t\). Since every reduced row of \(E\) is a Boolean union
of at most \(R\) vectors from \(\mathcal{B}\), choose one such representation
for each pick and press the corresponding treadles. The resulting thresholded
drawdown is exactly \(E\).

For \(R=1\), each reduced row must itself be one of the treadle vectors, so the
criterion reduces to counting distinct reduced row types. If pressing no treadle
is disallowed, one simply omits the empty union from
\(\mathcal{U}_R(\mathcal{B})\).

For a periodic prefabric, all rows and columns are determined by a finite
fundamental period block, so applying the same argument to that block gives the
periodic statement.
\end{proof}

Thus, when \(R>1\), the minimum number of treadles is no longer simply the
number of distinct reduced row types. It is the size of the smallest family of
basic treadle lift patterns whose Boolean unions of size at most \(R\) cover all
reduced row types.

The periodic version of this statement uses the language of prefabrics and
fabrics, which we recall here.

\begin{definition}[Prefabric]
A \emph{prefabric} is a formal infinite over-under assignment for two transverse
families of strands, one warp-like and one weft-like. A prefabric need not hang
together as a physical cloth; it may fall apart into independent layers, strips,
or components~\cite{grunbaum1980,clapham1980,enns1984}.
\end{definition}

\begin{definition}[Fabric and isonemal fabric]
A \emph{fabric} is a prefabric whose strands interlace sufficiently to hang
together. A fabric is \emph{isonemal} if its symmetry group acts transitively on
the set of strands; equivalently, for any two strands, there is a symmetry of
the fabric that sends one strand to the other~\cite{grunbaum1980,grunbaum1988}.
\end{definition}

The symmetry theory of periodic fabrics, including isonemal fabrics, is largely
orthogonal to shaft attainability. A design can be highly symmetric and still
require many shaft-column behaviors. Conversely, a design can be attainable on
few shafts while failing to be a good physical cloth because of floats,
instability, or other textile constraints.

The inverse problem considered here begins with a target drawdown \(D^\ast\) and
asks for shaft-loom data \((A,B,C)\) producing a drawdown
\[
  \widehat{D}_{ij}
  =
  \mathbf{1}\{(AB^T C)_{ij}>0\}
\]
that is equal to \(D^\ast\), or, when exact equality is impossible under the
available numbers of shafts, treadles, and simultaneously pressed treadles, is
close to \(D^\ast\) according to a chosen objective function.

This turns inverse drafting into a constrained binary approximation problem.
The exact problem asks whether \(D^\ast\) factors through a sufficiently small
tie-up matrix. The approximate problem asks which factorization should be chosen
when the target has too many distinct column or row behaviors. A natural first
objective is Hamming error,
\[
  \operatorname{err}(\widehat{D},D^\ast)
  =
  \#\{(i,j) : \widehat{D}_{ij} \ne D^\ast_{ij}\},
\]
but this is rarely the whole story. Different errors may have different visual
or structural importance. For sparse targets, moving an isolated black cell may
be preferable to toggling several neighboring cells. For line-like or motif-like
targets, preserving continuity and local shape may matter more than minimizing
raw binary disagreement. For cloth, objectives should also penalize long warp or
weft floats, disconnected prefabric behavior, or other non-clothlike structures.

The purpose of these notes is to develop and compare algorithms for this
approximate inversion problem. The central questions are:
\begin{enumerate}[label=(\roman*)]
  \item how to find exact factorizations when they exist;
  \item how to search efficiently when the target exceeds the available loom
  constraints;
  \item how different cost functions change the chosen approximation;
  \item how to incorporate clothlike constraints such as float tolerance; and
  \item how to test an inversion engine using targets known to be generated by
  random drafts as well as targets generated by random binary drawdowns.
\end{enumerate}

\section{Related Work}

There are two closely related mathematical traditions behind this note. The
first is the algebraic and algorithmic analysis of weave drafts. Lourie's
``loom-constrained designs'' paper formulates the problem of determining a loom
mechanism, threading-like initial conditions, and dynamic control information
from a binary representation of a weave~\cite{lourie1969}. Hoskins and Hoskins
studied matrix equations and factoring algorithms arising from structural
analysis of woven fabrics~\cite{hoskins1981,hoskins1983}, while later work on
binary interlacement arrays and bounded-float twills develops related finite
combinatorial models~\cite{hoskins1984,hpstreet1983}.

The second tradition is the geometry and symmetry theory of fabrics and
prefabrics. Gr\"unbaum and Shephard introduced the modern mathematical
treatment of satins, twills, prefabrics, and isonemal fabrics
\cite{grunbaum1980,grunbaum1985,grunbaum1986,grunbaum1988}. Roth classified the
symmetry groups of periodic isonemal fabrics~\cite{roth1993}, and Thomas
refined Roth's taxonomy for prefabrics with parallel axes, perpendicular axes,
and no axes of symmetry~\cite{thomas2009,thomas2010perp,thomas2010noaxes}.
Questions about whether a prefabric ``hangs together'' are separate from
symmetry classification and lead to their own graph-like connectivity
criteria~\cite{clapham1980,delaney1986,enns1984}.

The present notes use the same binary-matrix language, but focus on a slightly
different computational question: when an exact shaft-loom factorization is
unavailable or undesirable, how should one approximate the target drawdown? In
particular, we keep track of the integer redundancy matrix \(D=AB^TC\), allow
multiple treadles to be pressed on a pick, and compare objective functions that
may value visual locality, motif continuity, or clothlike float behavior
differently from raw Hamming error.

\section{Problem Setup}

Let \(D^\ast \in \{0,1\}^{P \times W}\) be a target drawdown. The approximate
inverse-drafting problem is to choose matrices
\[
  A \in \{0,1\}^{P \times T},
  \qquad
  B \in \{0,1\}^{S \times T},
  \qquad
  C \in \{0,1\}^{S \times W}
\]
so as to minimize a cost
\[
  C_{\mathrm{obj}}(\widehat{D},D^\ast;A,B,C),
  \qquad
  \widehat{D}_{ij}=\mathbf{1}\{(AB^T C)_{ij}>0\}.
\]
The matrix \(C\) is constrained to have exactly one \(1\) in each column, since
each warp end is threaded on exactly one shaft. The matrix \(A\) is constrained
by the maximum number of simultaneously pressed treadles:
\[
  \sum_{t=1}^{T} A_{it} \le R
  \qquad
  \text{for all } i=1,\dots,P.
\]

\begin{theorem}[Redundancy Matrix]
Let \(A\in\{0,1\}^{P\times T}\) be a treadling matrix, let
\(B\in\{0,1\}^{S\times T}\) be a tie-up matrix, and let
\(C\in\{0,1\}^{S\times W}\) be a threading matrix with exactly one nonzero entry
in each column. Define
\[
  D = A B^{T} C.
\]
Then \(D\in\mathbb{Z}_{\ge 0}^{P\times W}\), and \(D_{ij}\) is the number of
pressed treadles on pick \(i\) that lift the shaft containing warp end \(j\).
In particular, if
\[
  \widehat{D}_{ij}=\mathbf{1}\{D_{ij}>0\},
\]
then \(\widehat{D}_{ij}=1\) exactly when warp end \(j\) is lifted over pick
\(i\). Moreover, if at most \(R\) treadles are pressed on each pick, then
\[
  0 \le D_{ij} \le R
  \qquad
  \text{for all } i,j.
\]
\end{theorem}

\begin{proof}
By the definition of matrix multiplication,
\[
  D_{ij}
  =
  (AB^{T}C)_{ij}
  =
  \sum_{t=1}^{T}\sum_{s=1}^{S} A_{it}B_{st}C_{sj}.
\]
Since warp end \(j\) is threaded on exactly one shaft, there is a unique shaft
\(s(j)\) such that \(C_{s(j),j}=1\), and \(C_{sj}=0\) for \(s\ne s(j)\).
Therefore the double sum collapses to
\[
  D_{ij}
  =
  \sum_{t=1}^{T} A_{it}B_{s(j),t}.
\]
For a fixed pick \(i\), the factor \(A_{it}\) is \(1\) exactly when treadle
\(t\) is pressed. The factor \(B_{s(j),t}\) is \(1\) exactly when treadle \(t\)
lifts the shaft \(s(j)\) carrying warp end \(j\). Thus each summand is \(1\)
exactly for a pressed treadle that lifts the shaft containing warp end \(j\), and
is \(0\) otherwise. Hence \(D_{ij}\) counts those treadles.

It follows that \(D_{ij}>0\) exactly when at least one pressed treadle lifts the
shaft containing warp end \(j\). This is precisely the event that warp end \(j\)
is lifted over pick \(i\), so the binary drawdown is
\(\widehat{D}_{ij}=\mathbf{1}\{D_{ij}>0\}\). Finally, if at most \(R\) treadles
are pressed on pick \(i\), then at most \(R\) terms in
\(\sum_{t=1}^{T} A_{it}B_{s(j),t}\) can be nonzero, so \(D_{ij}\le R\).
\end{proof}

\clearpage
\begin{thebibliography}{99}

\bibitem{lourie1969}
Janice R. Lourie.
\newblock Loom-constrained designs: An algebraic solution.
\newblock In \emph{Proceedings of the 1969 24th National Conference}, ACM,
pages 185--192, 1969.

\bibitem{hoskins1981}
J. Hoskins and W. Hoskins.
\newblock The solutions of certain matrix equations arising from the structural
analysis of woven fabrics.
\newblock \emph{Ars Combinatoria}, 11:51--59, 1981.

\bibitem{hoskins1983}
J. Hoskins and W. Hoskins.
\newblock A faster algorithm for factoring binary matrices.
\newblock \emph{Ars Combinatoria}, 16-B:341--350, 1983.

\bibitem{hoskins1984}
J. Hoskins, A. Street, and R. Stanton.
\newblock Binary interlacement arrays and how to find them.
\newblock \emph{Congressus Numerantium}, 42:321--376, 1984.

\bibitem{hpstreet1983}
J. Hoskins, C. Praeger, and A. Street.
\newblock Twills with bounded float length.
\newblock \emph{Bulletin of the Australian Mathematical Society}, 28:255--281,
1983.

\bibitem{grunbaum1980}
Branko Gr\"unbaum and G. C. Shephard.
\newblock Satins and twills: An introduction to the geometry of fabrics.
\newblock \emph{Mathematics Magazine}, 53(3):139--161, 1980.

\bibitem{grunbaum1985}
Branko Gr\"unbaum and G. C. Shephard.
\newblock A catalogue of isonemal fabrics.
\newblock In J. E. Goodman, editor, \emph{Discrete Geometry and Convexity},
Annals of the New York Academy of Sciences, volume 440, pages 279--298, 1985.

\bibitem{grunbaum1986}
Branko Gr\"unbaum and G. C. Shephard.
\newblock An extension to the catalogue of isonemal fabrics.
\newblock \emph{Discrete Mathematics}, 60:155--192, 1986.

\bibitem{grunbaum1988}
Branko Gr\"unbaum and G. C. Shephard.
\newblock Isonemal fabrics.
\newblock \emph{The American Mathematical Monthly}, 95(1):5--30, 1988.

\bibitem{roth1993}
Richard L. Roth.
\newblock The symmetry groups of periodic isonemal fabrics.
\newblock \emph{Geometriae Dedicata}, 48:191--210, 1993.

\bibitem{thomas2009}
R. S. D. Thomas.
\newblock Isonemal prefabrics with only parallel axes of symmetry.
\newblock \emph{Discrete Mathematics}, 309(9):2696--2711, 2009.

\bibitem{thomas2010perp}
R. S. D. Thomas.
\newblock Isonemal prefabrics with perpendicular axes of symmetry.
\newblock \emph{Utilitas Mathematica}, 82:33--70, 2010.

\bibitem{thomas2010noaxes}
R. S. D. Thomas.
\newblock Isonemal prefabrics with no axes of symmetry.
\newblock \emph{Discrete Mathematics}, 310(8):1307--1324, 2010.

\bibitem{clapham1980}
C. R. J. Clapham.
\newblock When a fabric hangs together.
\newblock \emph{Bulletin of the London Mathematical Society}, 12:161--164,
1980.

\bibitem{delaney1986}
Cathy Delaney.
\newblock When a fabric hangs together.
\newblock \emph{Ars Combinatoria A}, 21:71--79, 1986.

\bibitem{enns1984}
T. C. Enns.
\newblock An efficient algorithm determining when a fabric hangs together.
\newblock \emph{Geometriae Dedicata}, 15:259--260, 1984.

\end{thebibliography}

\end{document}
